---
title: "Rosalind Strong Set — Solutions"
author: "ABI Omics"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

# Overview

Worked solutions for the Rosalind strong set. Several problems below lean on the
**apply family** (`sapply`, `vapply`, `lapply`, `mapply`, `apply`) instead of
explicit `for` loops — especially **DNA**, **HAMM**, **GC**, **CONS**, **PROT**,
and **SPLC**.

```{r helpers}
chars <- function(s) strsplit(s, "")[[1]]

# Small FASTA parser -> named character vector of sequences
parse_fasta <- function(text) {
  lines <- strsplit(text, "\n", fixed = TRUE)[[1]]
  lines <- lines[nzchar(trimws(lines))]
  header_idx <- which(startsWith(lines, ">"))
  ids <- sub("^>", "", lines[header_idx])
  # Build end index of each sequence block, then sapply to concatenate lines
  ends <- c(header_idx[-1] - 1L, length(lines))
  seqs <- sapply(seq_along(header_idx), function(i) {
    paste(lines[(header_idx[i] + 1L):ends[i]], collapse = "")
  })
  setNames(seqs, ids)
}

gc_content <- function(seq) {
  bases <- chars(seq)
  100 * mean(bases %in% c("G", "C"))
}
```

---

# DNA — Counting DNA Nucleotides

Count each base with `sapply` over `c("A","C","G","T")` instead of four separate
counters or a `for` loop.

```{r dna}
dna <- "AGCTTTTCATTCTGACTGCAACGGGCAATATGTCTCTGTGTGGATTAAAAAAAGAGTGTCTGATAGCAGC"

count_nucleotides <- function(dna) {
  bases <- chars(dna)
  counts <- sapply(c("A", "C", "G", "T"), function(b) sum(bases == b))
  # Rosalind wants space-separated A C G T
  paste(counts, collapse = " ")
}

count_nucleotides(dna)
# Expected: 20 12 17 21
```

(`table(factor(bases, levels = c("A","C","G","T")))` is also fine — same idea,
no loop.)

---

# RNA — Transcribing DNA into RNA

```{r rna}
dna <- "GATGGAACTTGACTACGTAAATT"

transcribe <- function(dna) gsub("T", "U", dna, fixed = TRUE)

transcribe(dna)
# Expected: GAUGGAACUUGACUACGUAAAUU
```

---

# REVC — Complementing a Strand of DNA

Map each base with a named vector, then reverse. `sapply`/`chars` keeps the
logic readable; `chartr` is a short alternative.

```{r revc}
dna <- "AAAACCCGGT"

rev_complement <- function(dna) {
  complement <- c(A = "T", T = "A", C = "G", G = "C")
  paste(rev(complement[chars(dna)]), collapse = "")
}

# Equivalent one-liner style:
# rev_complement <- function(dna) paste(rev(chars(chartr("ACGT", "TGCA", dna))), collapse = "")

rev_complement(dna)
# Expected: ACCGGGTTTT
```

---

# HAMM — Counting Point Mutations

Split both strings and compare vectors elementwise — no index-walking loop.
`sum(x != y)` is the natural R idiom; `mapply` makes the per-position idea
explicit if you want that shape.

```{r hamm}
s <- "GAGCCTACTAACGGGAT"
t <- "CATCGTAATGACGGCCT"

hamming <- function(s, t) {
  sum(chars(s) != chars(t))
}

# Explicit apply-family twin (same result):
# hamming <- function(s, t) sum(mapply(`!=`, chars(s), chars(t)))

hamming(s, t)
# Expected: 7
```

---

# GC — Computing GC Content

Parse FASTA once, then `sapply` a GC% function over every sequence and pick
`which.max`. That replace-a-loop pattern is the whole point of the apply family.

```{r gc}
gc_fasta <- ">Rosalind_6404
CCTGCGGAAGATCGGCACTAGAATAGCCAGAACCGTTTCTCTGAGGCTTCCGGCCTTCCC
TCCCACTAATAATTCTGAGG
>Rosalind_5959
CCATCGGTAGCGCATCCTTAGTCCAATTAAGTCCCTATCCAGGCGACGAAAGTTACAAG
>Rosalind_0808
CCACCCTCGTGGTATGGCTAGGCATTCAGGAACCGGAGAACGCTTCAGACCAGCCCGGAC
TGGGAACCTGCGGGCAGTAGGTGGAAT"

find_max_gc <- function(fasta_text) {
  seqs <- parse_fasta(fasta_text)
  gc_vals <- sapply(seqs, gc_content)
  winner <- names(which.max(gc_vals))
  cat(winner, "\n", sprintf("%.6f", gc_vals[winner]), sep = "")
  invisible(gc_vals)
}

find_max_gc(gc_fasta)
# Expected:
# Rosalind_0808
# 60.919540
```

---

# SUBS — Finding a Motif in DNA

Slide a window with `sapply` over candidate start positions; keep the hits
(1-based for Rosalind).

```{r subs}
s <- "GATATATGCATATACTT"
t <- "ATAT"

find_motif <- function(s, t) {
  n <- nchar(s) - nchar(t) + 1L
  hits <- which(sapply(seq_len(n), function(i) substr(s, i, i + nchar(t) - 1L) == t))
  paste(hits, collapse = " ")
}

find_motif(s, t)
# Expected: 2 4 10
```

(`gregexpr(t, s, fixed = TRUE)` also works; the `sapply` version makes the
sliding-window idea obvious.)

---

# CONS — Consensus and Profile

Build an $n \times L$ character matrix (`rbind` of split sequences), then
`apply(..., 2, ...)` down each **column** (position) to count bases and pick
the consensus letter. This is the textbook use of `apply` on a matrix.

```{r cons}
cons_fasta <- ">Rosalind_1
ATCCAGCT
>Rosalind_2
GGGCAACT
>Rosalind_3
ATGGATCT
>Rosalind_4
AAGCAACC
>Rosalind_5
TTGGAACT
>Rosalind_6
ATGCCATT
>Rosalind_7
ATGGCACT"

consensus_and_profile <- function(fasta_text) {
  seqs <- parse_fasta(fasta_text)
  mat <- do.call(rbind, lapply(seqs, chars))  # rows = sequences, cols = positions

  bases <- c("A", "C", "G", "T")
  # profile: each column -> counts in A/C/G/T order
  profile <- apply(mat, 2, function(col) {
    sapply(bases, function(b) sum(col == b))
  })
  rownames(profile) <- bases

  # consensus: most common base per column (ties -> first max, which is fine)
  consensus <- paste(apply(profile, 2, function(col) names(which.max(col))), collapse = "")

  cat(consensus, "\n", sep = "")
  for (b in bases) {
    cat(b, ": ", paste(profile[b, ], collapse = " "), "\n", sep = "")
  }
  invisible(list(consensus = consensus, profile = profile))
}

consensus_and_profile(cons_fasta)
# Expected consensus: ATGCAACT
# Profile:
# A: 5 1 0 0 5 5 0 0
# C: 0 0 1 4 2 0 6 1
# G: 1 1 6 3 0 1 0 0
# T: 1 5 0 0 0 1 1 6
```

---

# PROT — Translating RNA into Protein

Chop the RNA into length-3 codons with `substring` + `sapply`, look each up in
the codon table, then stop at the first `*`.

```{r prot}
rna <- "AUGGCCAUGGCGCCCAGAACUGAGAUCAAUAGUACCCGUAUUAACGGGUGA"

codon_table <- c(
  UUU="F", UUC="F", UUA="L", UUG="L",
  UCU="S", UCC="S", UCA="S", UCG="S",
  UAU="Y", UAC="Y", UAA="*", UAG="*",
  UGU="C", UGC="C", UGA="*", UGG="W",
  CUU="L", CUC="L", CUA="L", CUG="L",
  CCU="P", CCC="P", CCA="P", CCG="P",
  CAU="H", CAC="H", CAA="Q", CAG="Q",
  CGU="R", CGC="R", CGA="R", CGG="R",
  AUU="I", AUC="I", AUA="I", AUG="M",
  ACU="T", ACC="T", ACA="T", ACG="T",
  AAU="N", AAC="N", AAA="K", AAG="K",
  AGU="S", AGC="S", AGA="R", AGG="R",
  GUU="V", GUC="V", GUA="V", GUG="V",
  GCU="A", GCC="A", GCA="A", GCG="A",
  GAU="D", GAC="D", GAA="E", GAG="E",
  GGU="G", GGC="G", GGA="G", GGG="G"
)

translate_rna <- function(rna, codon_table) {
  n_codons <- nchar(rna) / 3
  starts <- seq(1, by = 3, length.out = n_codons)
  codons <- substring(rna, starts, starts + 2L)
  aas <- codon_table[codons]
  # Drop from the first stop codon onward
  stop_at <- match("*", aas)
  if (!is.na(stop_at)) aas <- aas[seq_len(stop_at - 1L)]
  paste(aas, collapse = "")
}

translate_rna(rna, codon_table)
# Expected: MAMAPRTEINSTRING
```

---

# Bonus: SPLC — RNA Splicing

Reuse earlier pieces. Remove each intron with `Reduce`/`lapply`, then
DNA → RNA → `translate_rna`. The apply family shows up in intron removal and in
translation.

```{r splc}
splc_fasta <- ">Rosalind_10
ATGGTCTACATAGCTGACAAACAGCACGTAGCAATCGGTCGAATCTCGAGAGGCATATGGTCACATGATCGGTCGAGCGTGTTTCAAAGTTTGCGCCTAG
>Rosalind_12
ATCGGTCGAA
>Rosalind_15
ATCGGTCGAGCGTGT"

splice_and_translate <- function(fasta_text, codon_table) {
  seqs <- parse_fasta(fasta_text)
  dna <- seqs[[1]]
  introns <- seqs[-1]

  # Peel off every intron (order does not matter for non-overlapping Rosalind cases)
  exons_dna <- Reduce(function(seq, intron) gsub(intron, "", seq, fixed = TRUE),
                      introns,
                      init = dna)

  rna <- transcribe(exons_dna)
  translate_rna(rna, codon_table)
}

splice_and_translate(splc_fasta, codon_table)
# Expected: MVYIADKQHVASREAYGHMFKVCA
```

If you prefer an explicit `sapply`/`lapply` shape for intron removal:

```{r splc-alt, eval=FALSE}
# Same idea, step-by-step with lapply:
remove_introns <- function(dna, introns) {
  Reduce(function(seq, intron) gsub(intron, "", seq, fixed = TRUE),
         as.list(introns),
         init = dna)
}
```

---

# Apply-family cheat sheet (for this set)

| Problem | Loop you might have written | Apply-shaped stand-in |
|---------|-----------------------------|------------------------|
| DNA | `for (b in bases) if (b=="A") a <- a+1 ...` | `sapply(c("A","C","G","T"), ...)` |
| HAMM | `for (i in seq_along(s)) if (s[i]!=t[i]) ...` | `sum(chars(s) != chars(t))` / `mapply` |
| GC | loop over FASTA records accumulating max | `sapply(seqs, gc_content)` then `which.max` |
| SUBS | loop start positions | `sapply(seq_len(n), ...)` + `which` |
| CONS | loop positions, then bases | `apply(mat, 2, ...)` profile; `lapply` to build matrix |
| PROT | loop `i in seq(1, n, by=3)` | `substring` starts + vectorized `codon_table[...]` |
| SPLC | loop removing each intron | `Reduce` over intron list, then reuse PROT |

RNA and REVC are short enough that a for-loop vs apply distinction barely matters;
the table above is where the apply habit pays off.
