---
title: "Rosalind Strong Set — Bioinformatics Problems in R"
author: "ABI Omics"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Overview

This set covers classical Rosalind DNA/RNA string problems. Work through them in order;
later problems reuse ideas from earlier ones.

**Strongly recommended problems**

| ID | Title |
|----|-------|
| DNA | Counting DNA Nucleotides |
| RNA | Transcribing DNA into RNA |
| REVC | Complementing a Strand of DNA |
| HAMM | Counting Point Mutations |
| GC | Computing GC Content |
| SUBS | Finding a Motif in DNA |
| CONS | Consensus and Profile |
| PROT | Translating RNA into Protein |

**Bonus (apply-family deep dive)**

| ID | Title |
|----|-------|
| SPLC | RNA Splicing |

> **Hint on style.** In several of these problems your first instinct may be a `for` loop.
> Before you reach for one, ask whether `sapply`, `lapply`, `vapply`, `mapply`, or
> `apply` can express the same idea more cleanly. The apply family is one of the
> clearest practical differences between R and Python — and it *is* useful here.

Helper ideas you may want (you do not have to use these names):

```{r helpers-sketch}
# Split a DNA/RNA string into a character vector of bases
chars <- function(s) strsplit(s, "")[[1]]

# Very small FASTA parser: returns a named character vector of sequences
# (concatenates multi-line sequence blocks)
parse_fasta <- function(text) {
  lines <- strsplit(text, "\n")[[1]]
  lines <- lines[nzchar(lines)]
  ids <- character()
  seqs <- character()
  current_id <- NULL
  current_seq <- character()
  flush <- function() {
    if (!is.null(current_id)) {
      ids <<- c(ids, current_id)
      seqs <<- c(seqs, paste(current_seq, collapse = ""))
    }
  }
  for (line in lines) {
    if (startsWith(line, ">")) {
      flush()
      current_id <- sub("^>", "", line)
      current_seq <- character()
    } else {
      current_seq <- c(current_seq, line)
    }
  }
  flush()
  setNames(seqs, ids)
}
```

For each problem: read the sample dataset, write your solution in the code chunk,
then check against the sample output. When you are ready, grab a dataset from
[Rosalind](https://rosalind.info) and submit.

---

# DNA — Counting DNA Nucleotides

**Problem.** Given a DNA string $s$ of length at most 1000 nt, return four integers
counting the respective number of times that the symbols **A**, **C**, **G**, and **T**
occur in $s$.

**Sample dataset**

```{r dna-data}
dna <- "AGCTTTTCATTCTGACTGCAACGGGCAATATGTCTCTGTGTGGATTAAAAAAAGAGTGTCTGATAGCAGC"
```

**Sample output:** `20 12 17 21`

```{r dna-solution}
# Your solution here.
# Prefer something apply-/vapply-shaped over an explicit for loop counting each base.
```

---

# RNA — Transcribing DNA into RNA

**Problem.** Given a DNA string $t$ corresponding to a coding strand, return its
transcribed RNA string $u$ (replace every **T** with **U**).

**Sample dataset**

```{r rna-data}
dna <- "GATGGAACTTGACTACGTAAATT"
```

**Sample output:** `GAUGGAACUUGACUACGUAAAUU`

```{r rna-solution}
# Your solution here.
```

---

# REVC — Complementing a Strand of DNA

**Problem.** Given a DNA string, return its reverse complement.
Complement: A↔T, C↔G. Then reverse the resulting string.

**Sample dataset**

```{r revc-data}
dna <- "AAAACCCGGT"
```

**Sample output:** `ACCGGGTTTT`

```{r revc-solution}
# Your solution here.
# One clean route: chars -> complement map -> reverse -> paste.
```

---

# HAMM — Counting Point Mutations

**Problem.** Given two DNA strings $s$ and $t$ of equal length, return the Hamming
distance $d_H(s, t)$ (number of differing positions).

**Sample dataset**

```{r hamm-data}
s <- "GAGCCTACTAACGGGAT"
t <- "CATCGTAATGACGGCCT"
```

**Sample output:** `7`

```{r hamm-solution}
# Your solution here.
# Vectorized comparison (or mapply) beats a for loop that walks both strings.
```

---

# GC — Computing GC Content

**Problem.** Given a multi-FASTA of DNA strings, return the ID of the string with
the highest GC-content, followed by its GC-content as a percentage.

GC-content of a string is $100 \times \frac{\#(G) + \#(C)}{\text{length}}$.

**Sample dataset**

```{r gc-data}
gc_fasta <- ">Rosalind_6404
CCTGCGGAAGATCGGCACTAGAATAGCCAGAACCGTTTCTCTGAGGCTTCCGGCCTTCCC
TCCCACTAATAATTCTGAGG
>Rosalind_5959
CCATCGGTAGCGCATCCTTAGTCCAATTAAGTCCCTATCCAGGCGACGAAAGTTACAAG
>Rosalind_0808
CCACCCTCGTGGTATGGCTAGGCATTCAGGAACCGGAGAACGCTTCAGACCAGCCCGGAC
TGGGAACCTGCGGGCAGTAGGTGGAAT"
```

**Sample output**

```
Rosalind_0808
60.919540
```

(Any precision comparable to this is fine for practice; Rosalind accepts rounding error.)

```{r gc-solution}
# Your solution here.
# Natural apply pattern: parse FASTA -> sapply each sequence for GC% -> which.max.
```

---

# SUBS — Finding a Motif in DNA

**Problem.** Given two DNA strings $s$ and $t$ (with $t$ a motif of $s$), return all
starting locations of $t$ in $s$. Locations use **1-based** indexing.

**Sample dataset**

```{r subs-data}
s <- "GATATATGCATATACTT"
t <- "ATAT"
```

**Sample output:** `2 4 10`

```{r subs-solution}
# Your solution here.
# Sliding window: sapply over start positions, keep those where the substring matches.
```

---

# CONS — Consensus and Profile

**Problem.** Given a collection of DNA strings of equal length in FASTA format,
return:

1. A consensus string (most frequent base at each position; ties may be broken any way)
2. The profile matrix (counts of A, C, G, T at each position), labeled rows

**Sample dataset**

```{r cons-data}
cons_fasta <- ">Rosalind_1
ATCCAGCT
>Rosalind_2
GGGCAACT
>Rosalind_3
ATGGATCT
>Rosalind_4
AAGCAACC
>Rosalind_5
TTGGAACT
>Rosalind_6
ATGCCATT
>Rosalind_7
ATGGCACT"
```

**Sample output** (one valid consensus; profile is unique)

```
ATGCAACT
A: 5 1 0 0 5 5 0 0
C: 0 0 1 4 2 0 6 1
G: 1 1 6 3 0 1 0 0
T: 1 5 0 0 0 1 1 6
```

```{r cons-solution}
# Your solution here.
# Think in matrices: split each sequence into chars, rbind into a matrix,
# then apply(..., 2, ...) down each column for the profile / consensus.
```

---

# PROT — Translating RNA into Protein

**Problem.** Given an RNA string corresponding to a protein-coding ORF (length
divisible by 3), translate it into its protein string using the standard genetic
code. Stop when you hit a stop codon; do **not** include the stop symbol.

**Sample dataset**

```{r prot-data}
rna <- "AUGGCCAUGGCGCCCAGAACUGAGAUCAAUAGUACCCGUAUUAACGGGUGA"
```

**Sample output:** `MAMAPRTEINSTRING`

Codon table (RNA → amino acid; `*` = stop):

```{r codon-table}
codon_table <- c(
  UUU="F", UUC="F", UUA="L", UUG="L",
  UCU="S", UCC="S", UCA="S", UCG="S",
  UAU="Y", UAC="Y", UAA="*", UAG="*",
  UGU="C", UGC="C", UGA="*", UGG="W",
  CUU="L", CUC="L", CUA="L", CUG="L",
  CCU="P", CCC="P", CCA="P", CCG="P",
  CAU="H", CAC="H", CAA="Q", CAG="Q",
  CGU="R", CGC="R", CGA="R", CGG="R",
  AUU="I", AUC="I", AUA="I", AUG="M",
  ACU="T", ACC="T", ACA="T", ACG="T",
  AAU="N", AAC="N", AAA="K", AAG="K",
  AGU="S", AGC="S", AGA="R", AGG="R",
  GUU="V", GUC="V", GUA="V", GUG="V",
  GCU="A", GCC="A", GCA="A", GCG="A",
  GAU="D", GAC="D", GAA="E", GAG="E",
  GGU="G", GGC="G", GGA="G", GGG="G"
)

translate_rna <- function(mRNA, RNA_prot_table=codon_table) {
        n_codons <- length(mRNA) / 3
        starts <- seq(1, by = 3, length.out = n_codons)
        codons <- substring(rna, starts, starts + 2L)
        aas <- codon_table[codons]
        # Drop from the first stop codon onward
        stop_at <- match("*", aas)
        if (!is.na(stop_at)) aas <- aas[seq_len(stop_at - 1L)]
        paste(aas, collapse = "")
}
```

```{r prot-solution}
# Your solution here.
# Slice the RNA into triplets, then sapply / vapply each codon through codon_table.
```

---

# Bonus: SPLC — RNA Splicing

**Problem.** Given a DNA string $s$ (coding strand) and a collection of substrings
of $s$ serving as **introns**, remove all introns from $s$, transcribe the remaining
exons to RNA, and translate to protein (as in PROT).

**Sample dataset**

```{r splc-data}
splc_fasta <- ">Rosalind_10
ATGGTCTACATAGCTGACAAACAGCACGTAGCAATCGGTCGAATCTCGAGAGGCATATGGTCACATGATCGGTCGAGCGTGTTTCAAAGTTTGCGCCTAG
>Rosalind_12
ATCGGTCGAA
>Rosalind_15
ATCGGTCGAGCGTGT"
```

**Sample output:** `MVYIADKQHVASREAYGHMFKVCA`

```{r splc-solution}
# Your solution here.
# Chain earlier pieces: remove introns (lapply/sapply), DNA→RNA, then PROT translation.
# This is a good place to reuse functions you already wrote above.
```

---

# Optional stretch

Once the sample datasets pass:

1. Download a Rosalind dataset for each problem and run your functions on it.
2. Refactor shared helpers (`chars`, `parse_fasta`, `translate_rna`, `gc_content`) into
   one place at the top of the document and reuse them.
3. For DNA / GC / CONS / PROT / HAMM / SPLC specifically, try rewriting any leftover
   `for` loops with apply-family calls and compare readability.
